# Reproducible Research: Peer Assessment 1
<br><br>

## Pre-Requisites

<br>

The following packages are required:

* data.tables
* xtable

the following R version was used to perform these calculations:
```{r, results='asis'}
xt <- version
#output results which include mean and median
print(xt, type='html')
```

## Loading and preprocessing the data

### Sources

<br>

The data for this assignment can be downloaded from the course web
site:

* Dataset: [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) [52K]

The variables included in this dataset are:

* **steps**: Number of steps taking in a 5-minute interval (missing
    values are coded as `NA`)
* **date**: The date on which the measurement was taken in YYYY-MM-DD
    format
* **interval**: Identifier for the 5-minute interval in which
    measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

### Preparation

The dataset used to perform this analysis was downloaded from source on 

Steps to download and prepare the file are as follows:

1. Create a temp. file name (eg tempfile())
2. Use download.file() to fetch the file into the temp. file
3. Use unz() to extract the target file from temp. file
4. Remove the temp file via unlink()

no further preparation is immediately performed, since the some output requires analysis of NA values.

The data used to prepare this document was downloaded on `r date()`

```{r}
  temp <- tempfile()
  download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",temp)
  data <- read.csv(unz(temp, "activity.csv"), header=T)
  unlink(temp)
```


## What is mean total number of steps taken per day?

For this part of the assignment,missing values in the dataset can be ignored.
histogram will igore `NA` values so no need to remove them, outputs are the same.

### Make a histogram of the total number of steps taken each day

First step here is to aggregate total steps for each day in the data set. We will use the a data table to speed up this process. In this case, we remove `NA` values from the sum.

```{r}
library(data.table)
dt<-data.table(data)
dt.agg <- dt[, sum(steps, na.rm=T), by=date]
hist(dt.agg$V1 , main="Histogram of Total Number of Steps /Day")
```
<br>

2. Calculate and report the **mean** and **median** total number of steps taken per day
<br>

```{r, results='asis'}
library(xtable)
# Transpose summary table to a format xtable can use
dt.sum <- summary(dt.agg$V1)
dt.t <- t(t(dt.sum))
xt <- xtable(dt.t)
#output results which include mean and median
print(xt, type='html')

```
<br>

## What is the average daily activity pattern?


1. Make a time series plot (i.e. `type = "l"`) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
#get rid of NA values for mean calcs
d<-na.omit(dt)
dt.ave <- d[, mean(steps, na.rm=T), by=interval]
setnames(dt.ave,c('interval','steps'))
plot(dt.ave$interval,dt.ave$steps, type="l")
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r, results='asis'}
maxInterval <- dt.ave[dt.ave$steps==max(dt.ave$steps)]
xt <- xtable(maxInterval)
#output results which include mean and median
print(xt, type='html')
```
<br>

Which indicates, the maximum interval is **`r  maxInterval[[1]]`** with an average of **`r  maxInterval[[2]]`** steps 

## Imputing missing values

### Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with `NA`s)

using the summary command, we see there are **`r paste(summary(data$steps)[7])`** `NA` values.

```{r}
summary(data)
```

### Devise a strategy for filling in all of the missing values in the dataset. 

Conventional Imputation typically uses multiple methods to substitutes estimated values for missing values

* Replacement with means
* Regression imputation (replace with conditional means)

The problem is that these methods 

* Often leads to biased parameter estimates (e.g., variances)
* Usually leads to standard error estimates that arebiased downward
  * Treats imputed data as real data, ignores inherent uncertainty in imputed values.

However, given that in this case the strategy does not need to be sophisticated. I have elected to replace all `NA` values with the mean value for that 5-minute interval. The dataset follows.

### Create a new dataset that is equal to the original dataset but with the missing data filled in.

The new dataset called `dt.impute` is created as follows:

1. Split the data set into two components, one with and one without `NA` values (`dt.noNA` and `dt.na` respectively)
2. merge the `dt.na` data set with the `dt.avg` set above, to get the average steps for the intervals in question
3. create a new dataset that replaces the NA values with the matched average step values
4. rbind the dt.noNA with the new dt.na data frame to create a complete table with no `NA` values

```{r}
dt.noNA <- data[!is.na(data$steps),]
dt.na <- data[is.na(data$steps),]
dt.m <- merge(x=dt.na, y=dt.ave, by='interval')
dt.impute <- rbind(dt.noNA, data.table(interval=dt.m$interval, steps=dt.m$steps.y, date=dt.m$date), use.names=T)
summary(dt.impute)
```

### Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.

```{r}
dt.aggImpute <- dt.impute[, sum(steps, na.rm=T), by=date]
hist(dt.aggImpute$V1 , main="Histogram of Total Number of Steps /Day")
```


## Are there differences in activity patterns between weekdays and weekends?
